---
title: "01 — Setup and Calibration"
author: "Natalie Love"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Overview

This notebook handles two things before you process any scans:

1. **Package installation** — installs `EBImage` (Bioconductor) and any other
   dependencies needed by the pipeline.
2. **Scale bar calibration** — verifies the pixel-to-millimeter conversion for
   your scanner so that all area measurements are in real-world units (cm²).

Run this notebook once before running `02_batch_segment.Rmd`. You only need to
re-run it if you change scanners or scanning resolution.

---

## 1. Install required packages

`EBImage` is a Bioconductor package designed for image processing and analysis.
It handles reading TIFFs, thresholding, connected-component labeling, and shape
feature extraction — everything this pipeline needs.

**Note:** The first installation can take several minutes because EBImage compiles
from source.

```{r install-packages, message=FALSE, warning=FALSE}
# Install BiocManager if not already present
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Install EBImage via Bioconductor
if (!require("EBImage", quietly = TRUE)) {
  BiocManager::install("EBImage", ask = FALSE)
}

# CRAN packages (tidyverse is almost certainly already installed)
cran_pkgs <- c("tidyverse", "fs")
for (pkg in cran_pkgs) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# Load everything
library(EBImage)
library(tidyverse)
library(fs)

cat("All packages loaded successfully.\n")
cat("EBImage version:", as.character(packageVersion("EBImage")), "\n")
```

---

## 2. User configuration

**Edit the paths and parameters in this chunk.** These values are used by all
three notebooks. Copy them into the matching chunk at the top of
`02_batch_segment.Rmd` and `03_sla_calculation.Rmd` as well.

```{r user-config}
# ── Path to the folder containing your raw TIFF scans ──────────────────────
# Use an absolute path or a path relative to the project root.
scan_folder <- "/Users/natalielove/Dropbox/1-Cloud/CBG/Research/Gymnocladus_dioicus/field_work/06232025_Ames_Iowa/leaf_scans"

# ── Output folders (created automatically if they don't exist) ──────────────
output_labels <- "output/labeled_images"   # annotated review PNGs
output_tables <- "output/area_tables"      # per-scan and master CSVs

# ── Scanner resolution ───────────────────────────────────────────────────────
# Dots per inch used when the scans were captured.
dpi <- 300

# ── Scale bar calibration ────────────────────────────────────────────────────
# Derived automatically from DPI:
#   300 pixels / 25.4 mm per inch = 11.811 pixels per mm
px_per_mm  <- dpi / 25.4
px_per_cm  <- px_per_mm * 10
px2_per_cm2 <- px_per_cm^2   # pixels² per cm²

cat(sprintf("Calibration: %.3f px/mm  |  %.3f px/cm  |  %.1f px²/cm²\n",
            px_per_mm, px_per_cm, px2_per_cm2))

# ── Segmentation thresholds ──────────────────────────────────────────────────
# Minimum object area to keep (cm²). Objects smaller than this are treated as
# dust, scanner noise, or tiny debris and discarded.
# Gymnocladus leaflets are typically > 1 cm², so 0.25 cm² is conservative.
min_area_cm2 <- 0.25

# Maximum elongation ratio (radius_max / radius_min) for an object to be kept.
# The white ruler is extremely elongated (long, thin bar). Leaf pieces and the
# color card are much more compact. A ratio above ~6 almost certainly indicates
# the ruler or a similar artifact.
max_elongation <- 6.0

# ── Output image resolution ─────────────────────────────────────────────────
# Labeled review images are saved at a fraction of the original resolution to
# keep file sizes manageable (originals are ~56 MB each).
# 0.25 = quarter resolution (saves 4x storage, still easy to read labels).
review_scale <- 0.25

# Print a summary of all key parameters
cat("\n── Configuration summary ──────────────────────────────────────────────\n")
cat("Scan folder:         ", scan_folder, "\n")
cat("Output labels:       ", output_labels, "\n")
cat("Output tables:       ", output_tables, "\n")
cat("DPI:                 ", dpi, "\n")
cat("Min object area:     ", min_area_cm2, "cm²\n")
cat("Max elongation ratio:", max_elongation, "\n")
cat("Review image scale:  ", review_scale, "\n")
```

---

## 3. Verify the scan folder and list files

```{r check-files}
# Create output directories if they don't exist
dir_create(output_labels)
dir_create(output_tables)

# List all TIFF files in the scan folder
tif_files <- dir_ls(scan_folder, regexp = "\\.tif$", ignore.case = TRUE)

cat(sprintf("Found %d TIFF files in:\n  %s\n\n", length(tif_files), scan_folder))
cat(paste(basename(tif_files), collapse = "\n"), "\n")
```

---

## 4. Load a test image and inspect it

We load a single scan to confirm image dimensions, confirm the file reads
correctly, and display a thumbnail so you can visually confirm the layout.

```{r load-test-image, fig.width=6, fig.height=9}
# Load the first scan as a test
test_path <- tif_files[1]
cat("Loading:", basename(test_path), "\n")

img <- readImage(test_path)

# Report basic properties
cat(sprintf(
  "Dimensions : %d x %d pixels (width x height)\n",
  dim(img)[1], dim(img)[2]
))
cat(sprintf(
  "Physical size at %d DPI: %.1f mm x %.1f mm\n",
  dpi,
  dim(img)[1] / px_per_mm,
  dim(img)[2] / px_per_mm
))
cat("Color channels:", dim(img)[3], "\n")
cat("Pixel value range: [", min(img), ",", max(img), "]\n")

# Display a thumbnail (resize to ~10% for fast rendering in the notebook)
display(resize(img, w = round(dim(img)[1] * 0.10)), method = "raster")
```

---

## 5. Scale bar calibration check

The pipeline derives pixels-per-mm directly from the scanner DPI (`300 / 25.4`).
This is mathematically exact *if* the scanner is truly set to 300 DPI.

Here we verify the calibration visually by cropping the bottom-right corner
of the test image (where the white ruler lives) and measuring how many pixels
correspond to a known distance on the ruler.

```{r calibration-check, fig.width=5, fig.height=3}
# Crop the bottom-right corner of the image where the ruler is located.
# Adjust these fractions if the ruler is positioned differently in your scans.
crop_bottom_frac <- 0.15   # take the bottom 15% of the image height
crop_right_frac  <- 0.35   # take the right 35% of the image width

img_w <- dim(img)[1]
img_h <- dim(img)[2]

x_start <- round(img_w * (1 - crop_right_frac))
y_start <- round(img_h * (1 - crop_bottom_frac))

ruler_crop <- img[x_start:img_w, y_start:img_h, ]

cat(sprintf("Ruler crop: %d x %d pixels\n", dim(ruler_crop)[1], dim(ruler_crop)[2]))
display(ruler_crop, method = "raster")
```

```{r expected-calibration}
# ── Expected pixel widths for common ruler distances at 300 DPI ─────────────
distances_mm <- c(10, 20, 30, 50, 100)
cat("Expected pixel widths at 300 DPI:\n")
for (d in distances_mm) {
  cat(sprintf("  %3d mm  =  %6.1f pixels\n", d, d * px_per_mm))
}

cat(sprintf(
  "\nTo verify: find two tick marks separated by a known distance on the ruler\n",
))
cat(sprintf(
  "in the cropped image above. The pixel distance between them should match\n"
))
cat(sprintf(
  "the values in the table. If they differ by more than ~2%%, update `dpi`\n"
))
cat(sprintf(
  "in the user configuration chunk above.\n"
))
```

---

## 6. Test segmentation on the example image

This section runs the full segmentation on the test image so you can visually
confirm that the algorithm correctly identifies the leaf pieces before running
the batch job.

See `02_batch_segment.Rmd` for full documentation of each step.

```{r test-segmentation, fig.width=6, fig.height=9, message=FALSE}
source_dir <- scan_folder   # used below for inline test

# ── Step 1: Convert to grayscale (luminance) ─────────────────────────────────
gray <- channel(img, "luminance")

# ── Step 2: Determine threshold using Otsu's method ─────────────────────────
# Otsu's method finds the threshold that minimises within-class variance between
# the dark background and the bright foreground (leaf material + ruler + card).
th <- otsu(gray)
cat(sprintf("Otsu threshold: %.4f  (pixel values range 0–1)\n", th))

# ── Step 3: Create binary mask ───────────────────────────────────────────────
binary <- gray > th

# ── Step 4: Morphological opening to remove isolated bright specks ───────────
# "Opening" = erosion followed by dilation. It removes objects smaller than the
# structuring element (a 3x3 square here) while preserving larger shapes.
binary_clean <- opening(binary, makeBrush(3, shape = "box"))

# ── Step 5: Fill interior holes ──────────────────────────────────────────────
# Leaf veins can create small dark holes inside the leaf mask. fillHull() fills
# any enclosed dark region within a bright object.
binary_filled <- fillHull(binary_clean)

# ── Step 6: Label connected components ──────────────────────────────────────
# Each spatially separate bright region gets a unique integer label.
labeled <- bwlabel(binary_filled)
n_raw <- max(labeled)
cat(sprintf("Connected components before filtering: %d\n", n_raw))

# ── Step 7: Compute shape features for each labeled object ──────────────────
features <- computeFeatures.shape(labeled)
# Columns: s.area, s.perimeter, s.radius.mean, s.radius.sd,
#          s.radius.max, s.radius.min

# ── Step 8: Filter by area ───────────────────────────────────────────────────
min_area_px <- min_area_cm2 * px2_per_cm2
keep_area   <- features[, "s.area"] >= min_area_px

# ── Step 9: Filter by elongation ─────────────────────────────────────────────
# elongation = longest radius / shortest radius
# A long, thin ruler will have a very large ratio; leaves will not.
elongation <- features[, "s.radius.max"] / pmax(features[, "s.radius.min"], 1)
keep_shape  <- elongation <= max_elongation

# Combine filters
keep        <- keep_area & keep_shape
valid_ids   <- which(keep)
cat(sprintf("Objects after filtering: %d\n", length(valid_ids)))

# ── Step 10: Get centroids for label placement ───────────────────────────────
moments <- computeFeatures.moment(labeled)

# ── Step 11: Display annotated review image ─────────────────────────────────
# Overlay outlines of surviving objects on the original image
img_annotated <- paintObjects(
  x      = labeled,
  tgt    = img,
  opac   = c(0.8, 0.3),          # (outline opacity, fill opacity)
  col    = c("#FF0000", "#FF000020"),   # red outline, faint red fill
  thick  = TRUE
)

display(resize(img_annotated, w = round(dim(img_annotated)[1] * review_scale)),
        method = "raster")
```

```{r test-area-table}
# Build a tidy area table for the test image
area_table <- tibble(
  scan_id     = tools::file_path_sans_ext(basename(test_path)),
  piece_num   = seq_along(valid_ids),
  orig_label  = valid_ids,
  area_px     = features[valid_ids, "s.area"],
  area_mm2    = area_px / (px_per_mm^2),
  area_cm2    = area_px / px2_per_cm2,
  centroid_x  = round(moments[valid_ids, "m.cx"]),
  centroid_y  = round(moments[valid_ids, "m.cy"]),
  elongation  = round(elongation[valid_ids], 2)
)

print(area_table)
cat(sprintf("\nTotal area of all pieces: %.2f cm²\n", sum(area_table$area_cm2)))
```

---

## 7. Summary

If the calibration check and test segmentation look correct, you are ready to
run `02_batch_segment.Rmd` on all scans.

Things to confirm before proceeding:

- [ ] The ruler crop shows a readable ruler with tick marks
- [ ] The expected pixel widths match what you see on the ruler
- [ ] The test segmentation correctly outlines the leaf pieces
- [ ] The number of detected objects is plausible for your scan
- [ ] No large non-leaf objects (ruler, color card) appear in the area table
      (if they do, increase `max_elongation` or adjust filters in 02)

```{r session-info}
sessionInfo()
```
