---
title: "01 — Setup and Calibration"
author: "Natalie Love"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Overview

This notebook handles two things before you process any scans:

1. **Package installation** — installs `EBImage` (Bioconductor) and any other
   dependencies needed by the pipeline.
2. **Scale bar calibration** — verifies the pixel-to-millimeter conversion for
   your scanner so that all area measurements are in real-world units (cm²).

Run this notebook once before running `02_batch_segment.Rmd`. You only need to
re-run it if you change scanners or scanning resolution.

---

## 1. Install required packages

`EBImage` is a Bioconductor package designed for image processing and analysis.
It handles reading TIFFs, thresholding, connected-component labeling, and shape
feature extraction — everything this pipeline needs.

**Note:** The first installation can take several minutes because EBImage compiles
from source.

```{r install-packages, message=FALSE, warning=FALSE}
# Install BiocManager if not already present
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Install EBImage via Bioconductor
if (!require("EBImage", quietly = TRUE)) {
  BiocManager::install("EBImage", ask = FALSE)
}

# CRAN packages (tidyverse is almost certainly already installed)
cran_pkgs <- c("tidyverse", "fs")
for (pkg in cran_pkgs) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# Load everything
library(EBImage)
library(tidyverse)
library(fs)

cat("All packages loaded successfully.\n")
cat("EBImage version:", as.character(packageVersion("EBImage")), "\n")
```

---

## 2. User configuration

**Edit the paths and parameters in this chunk.** These values are used by all
three notebooks. Copy them into the matching chunk at the top of
`02_batch_segment.Rmd` and `03_sla_calculation.Rmd` as well.

```{r user-config}
# ── Path to the folder containing your raw TIFF scans ──────────────────────
# Use an absolute path or a path relative to the project root.
scan_folder <- "data/leaf_scans"

# ── Output folders (created automatically if they don't exist) ──────────────
output_labels <- "output/labeled_images"   # annotated review PNGs
output_tables <- "output/area_tables"      # per-scan and master CSVs

# ── Scanner resolution ───────────────────────────────────────────────────────
# Dots per inch used when the scans were captured.
dpi <- 300

# ── Scale bar calibration ────────────────────────────────────────────────────
# Derived automatically from DPI:
#   300 pixels / 25.4 mm per inch = 11.811 pixels per mm
px_per_mm  <- dpi / 25.4
px_per_cm  <- px_per_mm * 10
px2_per_cm2 <- px_per_cm^2   # pixels² per cm²

cat(sprintf("Calibration: %.3f px/mm  |  %.3f px/cm  |  %.1f px²/cm²\n",
            px_per_mm, px_per_cm, px2_per_cm2))

# ── Segmentation thresholds ──────────────────────────────────────────────────
# Minimum object area to keep (cm²). Objects smaller than this are treated as
# dust, scanner noise, or tiny debris and discarded.
# Gymnocladus leaflets are typically > 1 cm², so 0.25 cm² is conservative.
min_area_cm2 <- 0.25

# Maximum elongation ratio (radius_max / radius_min) for an object to be kept.
# The white ruler is extremely elongated (long, thin bar). Leaf pieces and the
# color card are much more compact. A ratio above ~6 almost certainly indicates
# the ruler or a similar artifact.
max_elongation <- 6.0

# ── Output image resolution ─────────────────────────────────────────────────
# Labeled review images are saved at a fraction of the original resolution to
# keep file sizes manageable (originals are ~56 MB each).
# 0.25 = quarter resolution (saves 4x storage, still easy to read labels).
review_scale <- 0.25

# Print a summary of all key parameters
cat("\n── Configuration summary ──────────────────────────────────────────────\n")
cat("Scan folder:         ", scan_folder, "\n")
cat("Output labels:       ", output_labels, "\n")
cat("Output tables:       ", output_tables, "\n")
cat("DPI:                 ", dpi, "\n")
cat("Min object area:     ", min_area_cm2, "cm²\n")
cat("Max elongation ratio:", max_elongation, "\n")
cat("Review image scale:  ", review_scale, "\n")
```

---

## 3. Verify the scan folder and list files

```{r check-files}
# Create output directories if they don't exist
dir_create(output_labels)
dir_create(output_tables)

# List all TIFF files in the scan folder
tif_files <- dir_ls(scan_folder, regexp = "\\.tif$", ignore.case = TRUE)

cat(sprintf("Found %d TIFF files in:\n  %s\n\n", length(tif_files), scan_folder))
cat(paste(basename(tif_files), collapse = "\n"), "\n")
```

---

## 4. Load a test image and inspect it

We load a single scan to confirm image dimensions, confirm the file reads
correctly, and display a thumbnail so you can visually confirm the layout.

```{r load-test-image, fig.width=6, fig.height=9}
# Load the first scan as a test
test_path <- tif_files[1]
cat("Loading:", basename(test_path), "\n")

# all = FALSE reads only the first frame (the full-res scan).
# Scanner TIFFs often embed a lower-resolution thumbnail as a second frame;
# EBImage errors if it tries to stack frames of different dimensions.
img <- readImage(test_path, all = FALSE)

# Report basic properties
cat(sprintf(
  "Dimensions : %d x %d pixels (width x height)\n",
  dim(img)[1], dim(img)[2]
))
cat(sprintf(
  "Physical size at %d DPI: %.1f mm x %.1f mm\n",
  dpi,
  dim(img)[1] / px_per_mm,
  dim(img)[2] / px_per_mm
))
cat("Color channels:", dim(img)[3], "\n")
cat("Pixel value range: [", min(img), ",", max(img), "]\n")

# Display a thumbnail (resize to ~10% for fast rendering in the notebook)
display(resize(img, w = round(dim(img)[1] * 0.10)), method = "raster")
```

---

## 5. Scale bar calibration check

The pipeline derives pixels-per-mm directly from the scanner DPI (`300 / 25.4`).
This is mathematically exact *if* the scanner is truly set to 300 DPI.

Here we verify the calibration visually by cropping the bottom-right corner
of the test image (where the white ruler lives) and measuring how many pixels
correspond to a known distance on the ruler.

```{r calibration-check, fig.width=5, fig.height=3}
# Crop the bottom-right corner of the image where the ruler is located.
# Adjust these fractions if the ruler is positioned differently in your scans.
crop_bottom_frac <- 0.05   # take the bottom 15% of the image height
crop_right_frac  <- 0.55   # take the right 35% of the image width

img_w <- dim(img)[1]
img_h <- dim(img)[2]

x_start <- round(img_w * (1 - crop_right_frac))
y_start <- round(img_h * (1 - crop_bottom_frac))

ruler_crop <- img[x_start:img_w, y_start:img_h, ]

cat(sprintf("Ruler crop: %d x %d pixels\n", dim(ruler_crop)[1], dim(ruler_crop)[2]))
display(ruler_crop, method = "raster")
```

```{r verify-dpi-from-metadata}
# ── Read DPI directly from the TIFF file metadata ────────────────────────────
# Every TIFF produced by a flatbed scanner stores its resolution in the file
# header (XResolution / YResolution tags). Reading this is far more reliable
# than trying to measure pixel distances manually.
#
# The `tiff` package (installed automatically as an EBImage dependency) exposes
# these tags via readTIFF(..., info = TRUE).

library(tiff)

tif_raw  <- readTIFF(test_path, info = TRUE, all = FALSE)
file_dpi <- attr(tif_raw, "x.resolution")   # pixels per inch stored in file

if (is.null(file_dpi)) {
  cat("Resolution metadata not found in this TIFF. See manual fallback below.\n")
} else {
  cat(sprintf("DPI stored in TIFF metadata:  %.1f\n", file_dpi))
  cat(sprintf("DPI set in user-config chunk: %.1f\n", dpi))
  cat(sprintf("Difference:                   %.1f%%\n",
              abs(file_dpi - dpi) / dpi * 100))

  if (abs(file_dpi - dpi) / dpi > 0.02) {
    cat(sprintf(
      "\n>> Update `dpi <- %.0f` in the user-config chunk to match the file.\n",
      round(file_dpi)
    ))
  } else {
    cat("\nCalibration confirmed — config DPI matches file metadata.\n")
  }
}
```


### Optional manual fallback

```{r expected-calibration}
# ── Reference table: expected pixel widths at the configured DPI ─────────────
distances_mm <- c(10, 20, 30, 50, 100)
cat(sprintf("Expected pixel widths at %d DPI:\n", dpi))
for (d in distances_mm) {
  cat(sprintf("  %3d mm  =  %6.1f pixels\n", d, d * px_per_mm))
}
cat("\nYou can visually cross-check these values against the ruler crop above.\n")
```

If the TIFF metadata check above showed `NULL` (rare — some scanning software
does not write resolution tags), you can enter a measurement here instead.

To get a pixel measurement: save the ruler crop image, open it in **Preview**
(macOS), draw a selection over a known ruler distance using the Rectangle Select
tool, and read the width in pixels from the status bar at the bottom of the
window. Then fill in the two values below and run the chunk.

# confirmed DPI for Ames images on 03/01/2026

```{r manual-dpi-fallback, eval=FALSE}
# Only run this chunk if the metadata check above returned NULL.

# export image
# writeImage(ruler_crop, "output/scale_bar/scale_bar_Ames29376-01.tiff")

measured_dist_px <- 118   # <-- pixel width of your selection in Preview
known_mm         <- 10    # <-- real-world width of that same span on the ruler

measured_dpi <- (measured_dist_px / known_mm) * 25.4
cat(sprintf("Implied DPI from manual measurement: %.1f\n", measured_dpi))
cat(sprintf("DPI set in user-config chunk:        %.1f\n", dpi))

if (abs(measured_dpi - dpi) / dpi > 0.02) {
  cat(sprintf(
    "\n>> Update `dpi <- %.0f` in the user-config chunk and re-run.\n",
    round(measured_dpi)
  ))
} else {
  cat("\nCalibration looks good — within 2%%.\n")
}
```

---

## 6. Test segmentation on the example image

This section runs the full segmentation on the test image so you can visually
confirm that the algorithm correctly identifies the leaf pieces before running
the batch job.

See `02_batch_segment.Rmd` for full documentation of each step.

```{r test-segmentation, fig.width=6, fig.height=9, message=FALSE}
source_dir <- scan_folder   # used below for inline test

# ── Step 1: Convert to grayscale (luminance) ─────────────────────────────────
gray <- channel(img, "luminance")

# ── Step 2: Determine threshold using Otsu's method ─────────────────────────
# Otsu's method finds the threshold that minimises within-class variance between
# the dark background and the bright foreground (leaf material + ruler + card).
th <- otsu(gray)
cat(sprintf("Otsu threshold: %.4f  (pixel values range 0–1)\n", th))

# ── Step 3: Create binary mask ───────────────────────────────────────────────
binary <- gray > th

# ── Step 4: Morphological opening to remove isolated bright specks ───────────
# "Opening" = erosion followed by dilation. It removes objects smaller than the
# structuring element (a 3x3 square here) while preserving larger shapes.
binary_clean <- opening(binary, makeBrush(3, shape = "box"))

# ── Step 5: Fill interior holes ──────────────────────────────────────────────
# Leaf veins can create small dark holes inside the leaf mask. fillHull() fills
# any enclosed dark region within a bright object.
binary_filled <- fillHull(binary_clean)

# ── Step 6: Label connected components ──────────────────────────────────────
# Each spatially separate bright region gets a unique integer label.
labeled <- bwlabel(binary_filled)
n_raw <- max(labeled)
cat(sprintf("Connected components before filtering: %d\n", n_raw))

# ── Step 7: Compute shape features for each labeled object ──────────────────
features <- computeFeatures.shape(labeled)
# Columns: s.area, s.perimeter, s.radius.mean, s.radius.sd,
#          s.radius.max, s.radius.min

# ── Step 8: Filter by area ───────────────────────────────────────────────────
min_area_px <- min_area_cm2 * px2_per_cm2
keep_area   <- features[, "s.area"] >= min_area_px

# ── Step 9: Filter by elongation ─────────────────────────────────────────────
# elongation = longest radius / shortest radius
# A long, thin ruler will have a very large ratio; leaves will not.
elongation <- features[, "s.radius.max"] / pmax(features[, "s.radius.min"], 1)
keep_shape  <- elongation <= max_elongation

# Combine filters
keep        <- keep_area & keep_shape
valid_ids   <- which(keep)
cat(sprintf("Objects after filtering: %d\n", length(valid_ids)))

# ── Step 10: Get centroids for label placement ───────────────────────────────
moments <- computeFeatures.moment(labeled)

# ── Step 11: Display annotated review image with piece numbers ───────────────
# paintObjects() draws red outlines around each surviving object.
img_annotated <- paintObjects(
  x     = labeled,
  tgt   = img,
  opac  = c(1, 0),            # solid outline, no fill (cleaner for labels)
  col   = c("#FF3300", NA),
  thick = TRUE
)

# Resize for display
img_display <- resize(img_annotated, w = round(dim(img_annotated)[1] * review_scale))
out_w <- dim(img_display)[1]
out_h <- dim(img_display)[2]

# Scale centroid coordinates to match the resized image.
# m.cx / m.cy are in full-resolution pixel space; multiply by review_scale
# to get coordinates in the resized image space.
# Y must be flipped: image y increases downward, but R plots y increasing upward,
# so cy_plot = out_h - (m.cy * review_scale).
cx_plot <- moments[valid_ids, "m.cx"] * review_scale
cy_plot <- out_h - moments[valid_ids, "m.cy"] * review_scale

# Draw image using base R so we can overlay text with text()
par(mar = c(0, 0, 0, 0))
plot(NULL, xlim = c(0, out_w), ylim = c(0, out_h), asp = 1,
     axes = FALSE, xlab = "", ylab = "")
rasterImage(as.raster(img_display), 0, 0, out_w, out_h)

# Add piece numbers at each centroid
# Yellow with a dark outline gives visibility on both light and dark backgrounds
text(cx_plot, cy_plot,
     labels = seq_along(valid_ids),
     col    = "yellow",
     cex    = 1.8,
     font   = 2)   # bold
```

```{r test-area-table}
# Build a tidy area table for the test image
area_table <- tibble(
  scan_id     = tools::file_path_sans_ext(basename(test_path)),
  piece_num   = seq_along(valid_ids),
  orig_label  = valid_ids,
  area_px     = features[valid_ids, "s.area"],
  area_mm2    = area_px / (px_per_mm^2),
  area_cm2    = area_px / px2_per_cm2,
  centroid_x  = round(moments[valid_ids, "m.cx"]),
  centroid_y  = round(moments[valid_ids, "m.cy"]),
  elongation  = round(elongation[valid_ids], 2)
)

print(area_table)
cat(sprintf("\nTotal area of all pieces: %.2f cm²\n", sum(area_table$area_cm2)))
```

---

## 7. Summary

If the calibration check and test segmentation look correct, you are ready to
run `02_batch_segment.Rmd` on all scans.

Things to confirm before proceeding:

- [ ] The ruler crop shows a readable ruler with tick marks
- [ ] The expected pixel widths match what you see on the ruler
- [ ] The test segmentation correctly outlines the leaf pieces
- [ ] The number of detected objects is plausible for your scan
- [ ] No large non-leaf objects (ruler, color card) appear in the area table
      (if they do, increase `max_elongation` or adjust filters in 02)

```{r session-info}
sessionInfo()
```
