---
title: "02 — Batch Leaf Segmentation"
author: "Natalie Love"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Overview

This notebook processes every TIFF scan in your scan folder and produces two
outputs for each image:

1. **Labeled review PNG** — the original scan with each detected piece numbered
   in red so you can visually verify which pieces were found and spot any that
   should be excluded (e.g., the color card if it is not fully masked out).
2. **Per-scan area CSV** — one row per piece, with pixel area, area in mm², and
   area in cm².

At the end, all per-scan CSVs are combined into a single master table
(`master_leaf_areas.csv`) that feeds into `03_sla_calculation.Rmd`.

### Image analysis pipeline (one scan)

```
Read TIFF (RGB, 0–1)
  ↓ convert to grayscale (luminance)
  ↓ Otsu threshold → binary mask
  ↓ morphological opening (remove tiny bright specks / noise)
  ↓ fill enclosed holes (leaf veins create small dark holes)
  ↓ label connected components (bwlabel)
  ↓ compute shape features per object (area, radii)
  ↓ filter: drop objects < min_area_cm2
  ↓ filter: drop objects with elongation > max_elongation  ← removes ruler
  ↓ renumber surviving objects 1, 2, 3 …
  ↓ save annotated PNG + area CSV
```

---

## 1. Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(EBImage)
library(tidyverse)
library(fs)
```

---

## 2. User configuration

**Copy the same values you set in `01_setup_and_calibration.Rmd`.**

```{r user-config}
# ── Paths ────────────────────────────────────────────────────────────────────
scan_folder    <- "/Users/natalielove/Dropbox/1-Cloud/CBG/Research/Gymnocladus_dioicus/field_work/06232025_Ames_Iowa/leaf_scans"
output_labels  <- "output/labeled_images"
output_tables  <- "output/area_tables"

# ── Scanner calibration ───────────────────────────────────────────────────────
dpi            <- 300
px_per_mm      <- dpi / 25.4
px_per_cm      <- px_per_mm * 10
px2_per_cm2    <- px_per_cm^2

# ── Segmentation filters ──────────────────────────────────────────────────────
# Minimum object area to keep (cm²). Objects smaller than this are discarded
# as noise or debris. Gymnocladus leaflets are typically 2–10 cm², so 0.25 cm²
# is conservative enough to catch any fragmented rachis pieces.
min_area_cm2   <- 0.25

# Maximum elongation ratio (longest radius / shortest radius).
# The white ruler is extremely elongated — ratio typically 50–200.
# Gymnocladus leaflets range ~1.5–4. Rachis stems ~10–30.
# A cutoff of 10 keeps all leaflet shapes while removing rachis and ruler.
max_elongation <- 10.0

# ── Output image scale ────────────────────────────────────────────────────────
# Review PNGs are saved at this fraction of the original resolution.
# 0.25 = quarter size — legible for reviewing piece numbers, much smaller files.
review_scale   <- 0.25

# ── Create output directories if needed ──────────────────────────────────────
dir_create(output_labels)
dir_create(output_tables)

cat("Configuration loaded.\n")
```

---

## 3. Define the segmentation function

`segment_scan()` takes one TIFF path and the calibration/filter parameters,
and returns a named list with:

- `area_table` — tidy tibble with one row per detected piece
- `img_annotated` — EBImage Image object with numbered pieces overlaid on the
  original scan (for saving as a review PNG)
- `n_raw` — total connected components before filtering (diagnostic)
- `n_kept` — number of pieces after filtering

```{r define-function}
segment_scan <- function(tif_path,
                         px_per_mm      = 300 / 25.4,
                         min_area_cm2   = 0.25,
                         max_elongation = 10.0) {

  px2_per_cm2 <- (px_per_mm * 10)^2   # pixels² per cm²
  scan_id     <- tools::file_path_sans_ext(basename(tif_path))

  # ── 1. Read image ────────────────────────────────────────────────────────
  # all = FALSE reads only the first frame (the full-res scan).
  # Scanner TIFFs often embed a lower-resolution thumbnail as a second frame;
  # EBImage errors if it tries to stack frames of different dimensions.
  img <- readImage(tif_path, all = FALSE)
  img_w <- dim(img)[1]
  img_h <- dim(img)[2]

  # ── 2. Convert to grayscale (luminance) ──────────────────────────────────
  # Luminance = 0.2126 R + 0.7152 G + 0.0722 B (ITU-R BT.709 weights).
  # This is more perceptually accurate than a simple channel average and
  # weights green more heavily — useful since leaves are green.
  gray <- channel(img, "luminance")

  # ── 3. Otsu's threshold ───────────────────────────────────────────────────
  # Otsu's method automatically finds the threshold that minimises within-class
  # variance between the dark background and bright foreground. It adapts to
  # slight differences in scanner brightness across sessions.
  th     <- otsu(gray)
  binary <- gray > th

  # ── 4. Morphological opening (noise removal) ─────────────────────────────
  # Erosion removes any bright feature smaller than the 3×3 kernel; dilation
  # restores the edges of larger objects. Together this eliminates isolated
  # bright pixels (dust, scanner noise) without affecting leaf areas.
  binary_clean <- opening(binary, makeBrush(3, shape = "box"))

  # ── 5. Fill enclosed holes ────────────────────────────────────────────────
  # Leaf veins can appear slightly darker than the blade, sometimes creating
  # small "holes" in the binary mask. fillHull() fills any enclosed dark area
  # within a bright object, giving a solid leaf outline.
  binary_filled <- fillHull(binary_clean)

  # ── 6. Label connected components ────────────────────────────────────────
  # Each spatially separate bright region is assigned a unique integer ID.
  # Pieces that touch will be merged into one label — if this is a problem,
  # ensure leaves are well-separated on the scanner bed.
  labeled <- bwlabel(binary_filled)
  n_raw   <- max(labeled)

  # ── 7. Compute shape features ─────────────────────────────────────────────
  # computeFeatures.shape() returns a matrix (one row per label) with:
  #   s.area       — area in pixels
  #   s.perimeter  — perimeter in pixels
  #   s.radius.mean, s.radius.sd, s.radius.max, s.radius.min
  features <- computeFeatures.shape(labeled)

  # ── 8. Compute centroids ──────────────────────────────────────────────────
  # Used to place piece-number labels on the review image.
  # m.cx = x centroid (pixels from left edge, 1 to width)
  # m.cy = y centroid (pixels from top edge, 1 to height)
  moments <- computeFeatures.moment(labeled)

  # ── 9. Filter: minimum area ───────────────────────────────────────────────
  min_area_px <- min_area_cm2 * px2_per_cm2
  keep_area   <- features[, "s.area"] >= min_area_px

  # ── 10. Filter: maximum elongation (removes ruler) ────────────────────────
  # Elongation = longest principal radius / shortest principal radius.
  # pmax(..., 1) prevents division by zero for degenerate single-pixel objects.
  elongation  <- features[, "s.radius.max"] / pmax(features[, "s.radius.min"], 1)
  keep_shape  <- elongation <= max_elongation

  valid_ids <- which(keep_area & keep_shape)
  n_kept    <- length(valid_ids)

  # ── 11. Build area table ──────────────────────────────────────────────────
  # piece_num is a sequential integer assigned to surviving objects (1, 2, …).
  # orig_label is the raw EBImage label ID (useful for debugging if a piece
  # looks wrong in the review image).
  area_table <- tibble(
    scan_id    = scan_id,
    piece_num  = seq_along(valid_ids),
    orig_label = valid_ids,
    area_px    = features[valid_ids, "s.area"],
    area_mm2   = round(area_px / (px_per_mm^2), 3),
    area_cm2   = round(area_px / px2_per_cm2, 4),
    centroid_x = round(moments[valid_ids, "m.cx"]),
    centroid_y = round(moments[valid_ids, "m.cy"]),
    elongation = round(elongation[valid_ids], 2)
  )

  # ── 12. Build annotated review image ─────────────────────────────────────
  # paintObjects() draws a red outline around each surviving labeled object on
  # a copy of the original image. We then annotate with piece numbers (see
  # the batch loop below, which handles text rendering via R graphics).
  #
  # To highlight only valid objects, zero out non-valid labels in a copy of
  # the labeled mask, then paint.
  labeled_valid        <- labeled
  all_ids              <- seq_len(n_raw)
  invalid_ids          <- setdiff(all_ids, valid_ids)
  for (bad_id in invalid_ids) {
    labeled_valid[labeled_valid == bad_id] <- 0L
  }

  img_annotated <- paintObjects(
    x    = labeled_valid,
    tgt  = img,
    opac = c(1, 0),              # solid outline, no fill
    col  = c("#FF3300", NA),     # red outlines only
    thick = TRUE
  )

  list(
    area_table    = area_table,
    img_annotated = img_annotated,
    moments       = moments,
    valid_ids     = valid_ids,
    n_raw         = n_raw,
    n_kept        = n_kept
  )
}
```

---

## 4. Batch process all scans

This loop calls `segment_scan()` on every TIFF in the scan folder, saves
outputs, and appends results to a running list.

A progress counter keeps you informed, and each scan gets a one-line summary.

```{r batch-process, message=FALSE, warning=FALSE}
# Collect all TIFF paths
tif_files <- dir_ls(scan_folder, regexp = "\\.tif$", ignore.case = TRUE)
n_scans   <- length(tif_files)
cat(sprintf("Processing %d scans...\n\n", n_scans))

# Container for all per-scan results
all_areas <- vector("list", n_scans)

for (i in seq_along(tif_files)) {

  tif_path <- tif_files[i]
  scan_id  <- tools::file_path_sans_ext(basename(tif_path))

  cat(sprintf("[%d/%d] %s ...", i, n_scans, scan_id))

  # ── Run segmentation ────────────────────────────────────────────────────
  result <- segment_scan(
    tif_path      = tif_path,
    px_per_mm     = px_per_mm,
    min_area_cm2  = min_area_cm2,
    max_elongation = max_elongation
  )

  cat(sprintf(" %d pieces found (from %d components)\n",
              result$n_kept, result$n_raw))

  # ── Save per-scan area CSV ───────────────────────────────────────────────
  csv_path <- file.path(output_tables, paste0(scan_id, "_areas.csv"))
  write_csv(result$area_table, csv_path)

  # ── Save labeled review PNG ──────────────────────────────────────────────
  # The PNG is saved at review_scale of original resolution to reduce disk use.
  # We open a PNG device, draw the annotated image, then add red text labels
  # at each piece centroid.
  png_path <- file.path(output_labels, paste0(scan_id, "_labeled.png"))

  img_ann  <- result$img_annotated
  out_w    <- round(dim(img_ann)[1] * review_scale)
  out_h    <- round(dim(img_ann)[2] * review_scale)

  # Scale factor for centroid coordinates and text size
  scale_cx <- out_w / dim(img_ann)[1]
  scale_cy <- out_h / dim(img_ann)[2]

  png(png_path, width = out_w, height = out_h, units = "px", bg = "black")
  par(mar = c(0, 0, 0, 0), oma = c(0, 0, 0, 0))

  # Render image as a raster within a [0,1] x [0,1] plot space
  plot(
    NULL,
    xlim = c(0, 1), ylim = c(0, 1),
    xlab = "", ylab = "",
    axes = FALSE, asp = out_h / out_w
  )
  rasterImage(as.raster(img_ann), 0, 0, 1, 1)

  # Overlay piece numbers at each centroid
  # Centroid coordinates from EBImage: m.cx (x from left), m.cy (y from top).
  # In R plot space: x_norm = m.cx / img_width, y_norm = 1 - m.cy / img_height
  if (result$n_kept > 0) {
    cx_norm <- result$moments[result$valid_ids, "m.cx"] / dim(img_ann)[1]
    cy_norm <- 1 - result$moments[result$valid_ids, "m.cy"] / dim(img_ann)[2]

    text(
      x      = cx_norm,
      y      = cy_norm,
      labels = seq_along(result$valid_ids),
      col    = "red",
      cex    = 2.2,
      font   = 2    # bold
    )
  }
  dev.off()

  # ── Store results ────────────────────────────────────────────────────────
  all_areas[[i]] <- result$area_table
}

cat("\nDone! All scans processed.\n")
```

---

## 5. Compile master area table

Combine all per-scan CSVs into one file. This is the primary input for
`03_sla_calculation.Rmd`.

```{r compile-master}
master_areas <- bind_rows(all_areas)

# Add individual_id by stripping the trailing scan number (-01, -02, -03)
# from the scan filename. The individual identifier is everything before
# the last hyphen+digits suffix.
# Example: "Ames29376-01_A27-02" → individual "Ames29376-01_A27"
master_areas <- master_areas %>%
  mutate(individual_id = str_replace(scan_id, "-\\d{2}$", ""))

# Save master CSV
master_path <- file.path(output_tables, "master_leaf_areas.csv")
write_csv(master_areas, master_path)

cat(sprintf(
  "Master table saved to: %s\n%d rows | %d scans | %d individuals\n",
  master_path,
  nrow(master_areas),
  n_distinct(master_areas$scan_id),
  n_distinct(master_areas$individual_id)
))
```

---

## 6. Diagnostic summary

Quick checks to catch any scans with unexpectedly high or low piece counts,
which could indicate a segmentation problem.

```{r diagnostics}
# Pieces per scan
pieces_per_scan <- master_areas %>%
  count(scan_id, name = "n_pieces") %>%
  arrange(n_pieces)

cat("── Pieces per scan ─────────────────────────────────────────────────────\n")
print(pieces_per_scan, n = Inf)

# Flag any scans with very few or very many pieces
expected_min <- 3    # a scan with < 3 pieces is probably incomplete
expected_max <- 40   # > 40 is unusual for a compound Gymnocladus leaf

flagged <- pieces_per_scan %>%
  filter(n_pieces < expected_min | n_pieces > expected_max)

if (nrow(flagged) > 0) {
  cat("\n⚠ Scans with unexpected piece counts — review their labeled PNGs:\n")
  print(flagged)
} else {
  cat("\nAll scans within expected piece count range.\n")
}
```

```{r area-summary}
# Area distribution across all pieces
cat("\n── Area distribution (cm²) ────────────────────────────────────────────\n")
summary(master_areas$area_cm2)

# Histogram
ggplot(master_areas, aes(x = area_cm2)) +
  geom_histogram(bins = 40, fill = "#2d8e4a", color = "white") +
  labs(
    title = "Distribution of individual piece areas",
    x = "Area (cm²)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r total-per-individual}
# Total area per individual (summing all pieces across all scans)
# In 03_sla_calculation.Rmd you can choose to exclude specific piece IDs before
# summing. This is just a preview using all detected pieces.
total_per_ind <- master_areas %>%
  group_by(individual_id) %>%
  summarise(
    n_scans      = n_distinct(scan_id),
    n_pieces     = n(),
    total_area_cm2 = round(sum(area_cm2), 3),
    .groups = "drop"
  ) %>%
  arrange(individual_id)

cat("── Total leaf area per individual (all pieces, preview) ───────────────\n")
print(total_per_ind, n = Inf)
```

---

## 7. How to review labeled images

The review PNGs in `output/labeled_images/` show each detected piece outlined
in red and numbered. Open them in any image viewer.

**If you see non-leaf objects with a piece number** (e.g., part of the color
calibration card that was not masked out by the elongation filter), note those
piece numbers. In `03_sla_calculation.Rmd` you can provide an exclusion table
that drops specific piece IDs from specific scans before summing.

**If a leaf piece is missing** (not outlined), it likely fell below the
`min_area_cm2` threshold or was too dark to threshold correctly. Check whether
the leaf was fully flat against the scanner glass.

**If two leaf pieces appear merged into one**, they were touching or overlapping
on the scanner bed. Overlapping pieces cannot be automatically separated without
additional watershed processing. The best solution is to rescan with the pieces
clearly separated.

```{r session-info}
sessionInfo()
```
