---
title: "03 — SLA Calculation"
author: "Natalie Love"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Overview

**Specific Leaf Area (SLA)** = leaf area (cm²) / dry mass (g)

High SLA → thin, cheap-to-build leaves typical of fast-growing, shade-tolerant
or resource-rich environments.
Low SLA → thick, dense leaves typical of drought-tolerant or high-light
environments.

This notebook:

1. Loads the master leaf-area table produced by `02_batch_segment.Rmd`
2. Lets you optionally exclude specific piece IDs (e.g., color card fragments
   that were not automatically filtered)
3. Loads your dry-weight data and joins it to the area data by individual ID
4. Calculates total leaf area per individual (summing across all scans) and SLA
5. Exports the final SLA table and diagnostic plots

---

## 1. Load packages

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
```

---

## 2. User configuration

```{r user-config}
# ── Input files ───────────────────────────────────────────────────────────────
# Master area table produced by 02_batch_segment.Rmd
master_areas_path <- "output/area_tables/master_leaf_areas.csv"

# Your dry-weight data CSV. It must contain at least two columns:
#   - an individual ID column that matches the 'individual_id' field in the
#     master area table (e.g., "Ames29376-01_A27")
#   - a dry weight column in grams
#
# Set the column names to match your file:
dw_path          <- ""       # path to your dry weight CSV — FILL THIS IN
dw_id_col        <- "individual_id"   # column name for the individual ID
dw_weight_col    <- "dry_weight_g"    # column name for dry weight (grams)

# ── Output ────────────────────────────────────────────────────────────────────
output_tables    <- "output/area_tables"
sla_output_path  <- file.path(output_tables, "sla_results.csv")

# ── Exclusions ────────────────────────────────────────────────────────────────
# After reviewing the labeled PNGs in output/labeled_images/, you may find that
# certain piece numbers in certain scans represent the color calibration card or
# other non-leaf material that was not automatically removed.
#
# Specify those here as a data frame with two columns:
#   scan_id   — the scan filename without .tif extension
#   piece_num — the piece number shown in red on the labeled PNG
#
# Example:
#   exclusions <- tribble(
#     ~scan_id,              ~piece_num,
#     "Ames29376-01_A27-01",  5,
#     "Ames32901-07_E61-02",  1
#   )
#
# Leave as an empty tibble if no exclusions are needed:
exclusions <- tibble(scan_id = character(), piece_num = integer())

# ── Area summation scope ──────────────────────────────────────────────────────
# Choose which scans to include when summing area per individual.
# This is useful if, for example, one scan was re-done and you want to exclude
# the original, or if you only have dry weights for certain scans.
#
# Set to NULL to include all scans; or provide a character vector of scan_ids
# to include (everything else is excluded from the sum):
include_scans <- NULL   # e.g., c("Ames29376-01_A27-01", "Ames29376-01_A27-02")
```

---

## 3. Load and filter the master area table

```{r load-areas}
master_areas <- read_csv(master_areas_path, show_col_types = FALSE)

cat(sprintf(
  "Master areas: %d pieces | %d scans | %d individuals\n",
  nrow(master_areas),
  n_distinct(master_areas$scan_id),
  n_distinct(master_areas$individual_id)
))

glimpse(master_areas)
```

```{r apply-exclusions}
# Remove any pieces flagged in the exclusions table above
if (nrow(exclusions) > 0) {
  before <- nrow(master_areas)
  master_areas <- master_areas %>%
    anti_join(exclusions, by = c("scan_id", "piece_num"))
  after <- nrow(master_areas)
  cat(sprintf("Excluded %d pieces based on manual exclusion list.\n",
              before - after))
} else {
  cat("No manual exclusions applied.\n")
}
```

```{r apply-scan-filter}
# Optionally restrict to a specified set of scans
if (!is.null(include_scans)) {
  before <- nrow(master_areas)
  master_areas <- master_areas %>%
    filter(scan_id %in% include_scans)
  after <- nrow(master_areas)
  cat(sprintf("Restricted to %d specified scans. Removed %d rows.\n",
              length(include_scans), before - after))
} else {
  cat("Using all scans.\n")
}
```

---

## 4. Sum area per individual

Each individual may have multiple scans (e.g., 3 batches of leaflets scanned
separately). Summing across all scans gives the total leaf area weighed for
that individual.

```{r sum-per-individual}
area_per_individual <- master_areas %>%
  group_by(individual_id) %>%
  summarise(
    n_scans        = n_distinct(scan_id),
    n_pieces       = n(),
    total_area_cm2 = round(sum(area_cm2), 4),
    .groups        = "drop"
  ) %>%
  arrange(individual_id)

cat("── Total leaf area per individual ──────────────────────────────────────\n")
print(area_per_individual, n = Inf)
```

---

## 5. Load dry weight data and join

```{r load-dry-weights}
# ── Guard: check the dry weight path is set ───────────────────────────────────
if (dw_path == "") {
  message(
    "⚠  No dry weight file specified. Set `dw_path` in the user configuration\n",
    "   chunk above and re-run this notebook.\n\n",
    "   The area-per-individual table has been calculated and is available as\n",
    "   `area_per_individual`. SLA cannot be computed without dry weights."
  )
  knitr::knit_exit()
}

dry_weights <- read_csv(dw_path, show_col_types = FALSE)

# Rename columns to standard names for the join
dry_weights <- dry_weights %>%
  rename(
    individual_id = all_of(dw_id_col),
    dry_weight_g  = all_of(dw_weight_col)
  ) %>%
  select(individual_id, dry_weight_g)

cat(sprintf("Dry weights loaded: %d individuals\n", nrow(dry_weights)))
```

```{r join-areas-weights}
# Left join: keep all individuals with area data; flag those missing dry weight
sla_data <- area_per_individual %>%
  left_join(dry_weights, by = "individual_id")

# Report any individuals without a dry weight match
missing_dw <- sla_data %>% filter(is.na(dry_weight_g))
if (nrow(missing_dw) > 0) {
  cat(sprintf(
    "⚠  %d individuals have area data but no dry weight:\n", nrow(missing_dw)
  ))
  print(missing_dw$individual_id)
}

# Report any dry weights without matching area data
missing_area <- dry_weights %>%
  anti_join(area_per_individual, by = "individual_id")
if (nrow(missing_area) > 0) {
  cat(sprintf(
    "⚠  %d individuals have dry weight but no area data:\n", nrow(missing_area)
  ))
  print(missing_area$individual_id)
}
```

---

## 6. Calculate SLA

SLA is expressed here in two common units:

- **cm²/g** — the raw ratio from this analysis
- **m²/kg** — the standard unit in the TRY Plant Trait Database and most
  literature (numerically equivalent to cm²/g, since both ×/÷ by 10,000)

```{r calculate-sla}
sla_results <- sla_data %>%
  mutate(
    sla_cm2_per_g  = round(total_area_cm2 / dry_weight_g, 2),
    sla_m2_per_kg  = sla_cm2_per_g   # cm²/g is numerically equal to m²/kg
  ) %>%
  filter(!is.na(sla_cm2_per_g))  # drop rows where dry weight was missing

cat("── SLA results ─────────────────────────────────────────────────────────\n")
print(sla_results %>% select(individual_id, total_area_cm2, dry_weight_g,
                              sla_cm2_per_g, sla_m2_per_kg),
      n = Inf)
```

---

## 7. Export results

```{r export-results}
write_csv(sla_results, sla_output_path)
cat(sprintf("SLA results saved to: %s\n", sla_output_path))

# Also save the area-per-individual table in case dry weights come later
write_csv(area_per_individual,
          file.path(output_tables, "area_per_individual.csv"))
cat("Area-per-individual table also saved.\n")
```

---

## 8. Summary statistics and plots

```{r sla-summary}
cat("── SLA summary (cm²/g = m²/kg) ─────────────────────────────────────────\n")
summary(sla_results$sla_cm2_per_g)
cat(sprintf("n = %d individuals\n", nrow(sla_results)))
```

```{r sla-histogram, fig.width=7, fig.height=4}
ggplot(sla_results, aes(x = sla_cm2_per_g)) +
  geom_histogram(bins = 20, fill = "#2d8e4a", color = "white") +
  geom_vline(aes(xintercept = median(sla_cm2_per_g, na.rm = TRUE)),
             linetype = "dashed", color = "grey30") +
  labs(
    title = "Distribution of SLA across individuals",
    subtitle = paste0("Median = ",
                      round(median(sla_results$sla_cm2_per_g, na.rm = TRUE), 1),
                      " cm²/g"),
    x = "SLA (cm²/g)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r area-vs-weight, fig.width=6, fig.height=5}
ggplot(sla_results, aes(x = dry_weight_g, y = total_area_cm2)) +
  geom_point(color = "#2d8e4a", alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linewidth = 0.8) +
  labs(
    title  = "Total leaf area vs. dry weight",
    x      = "Dry weight (g)",
    y      = "Total leaf area (cm²)"
  ) +
  theme_minimal()
```

---

## 9. Extracting piece-level data for flexible area sums

The `master_leaf_areas.csv` table retains one row per piece, so you can always
re-sum in different ways — for example, to include only leaflets and exclude
the rachis, or to sum only scans 1 and 2 for a given individual.

```{r flexible-sum-example, eval=FALSE}
# Example: sum only pieces with area_cm2 > 1 cm² (likely only true leaflets)
leaflets_only <- master_areas %>%
  filter(area_cm2 > 1) %>%
  group_by(individual_id) %>%
  summarise(total_leaflet_area_cm2 = sum(area_cm2), .groups = "drop")

# Example: sum scans -01 and -02 only (exclude third scan batch)
two_scans <- master_areas %>%
  filter(str_detect(scan_id, "-(01|02)$")) %>%
  group_by(individual_id) %>%
  summarise(total_area_cm2 = sum(area_cm2), .groups = "drop")
```

---

## 10. Field column descriptions

| Column | Units | Description |
|---|---|---|
| `individual_id` | — | Tree identifier (`accession_individual`, e.g. `Ames29376-01_A27`) |
| `n_scans` | — | Number of separate scan files included in the sum |
| `n_pieces` | — | Total number of detected leaf pieces across all scans |
| `total_area_cm2` | cm² | Sum of all piece areas |
| `dry_weight_g` | g | Oven-dry mass of the sampled material |
| `sla_cm2_per_g` | cm²/g | Specific leaf area |
| `sla_m2_per_kg` | m²/kg | Same value, standard TRY database units |

```{r session-info}
sessionInfo()
```
